Add Plant Functional Type (PFT) labels

### 2021-04-22 R/add_plant_functional_type.Rmd
up

```{r message=FALSE}
library(tidyverse)
library(here)
library(dplyr)
library(kableExtra)
```

Plant functional types (PFTs) of interest: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.

```{r load_cleaned_spectra, message=FALSE}
# read cleaned spectral library file 
cleaned_spectra <- read_csv(here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
```

@@

```{r}
library(neonUtilities)
```

```{r}
# download woody vegetation structure data for all sites and years 
out_dir <- here::here("analysis","data","raw_data")

@@

```{r}
# read the vst_apparentindividual table, which has growthForm data
vst_apparentindividual <- read_csv(out_dir <- here::here("analysis","data","raw_data", "filesToStack10098",
                                                         "stackedFiles", "vst_apparentindividual.csv"))

@@

# simplify the growth form into "tree", "shrub", "saplng" etc.  
cleaned_spectra_merged <- cleaned_spectra_merged %>% 
  rowwise() %>%
  # parse out the last word in each entry's growthForm
  # make this code more efficient????? 
  mutate(growthForm_simple = rev(strsplit(growthForm, " ")[[1]])[1])

@@

```{r}
# summarize the merged growthform data
cleaned_spectra_merged %>% 
  distinct(spectraID, .keep_all = TRUE) %>% 
  group_by(genusSpecies, growthForm_simple) %>% 
    tally() 
```

@@

```{r}
# list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
# PFTs of interest: coniferous_forest, deciduous_forest, woody_shrub, grass_herb, and bare_ground.
taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
                               commonName = NA,
                               pft = NA) %>% 
  # arrange alphabetically
  dplyr::arrange(genusSpecies)

@@

  if(genusSpecies == "Abies amabilis"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "Pacific silver fir"}

  else if(genusSpecies == "Abies balsamea"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "balsam fir"}

  else if(genusSpecies == "Abies lasiocarpa"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$commonName[i] = "subalpine fir"}

  else if(genusSpecies == "Acacia greggii"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "catclaw acacia"}

  else if(genusSpecies == "Acer barbatum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "Southern sugar maple"}

  else if(genusSpecies == "Acer circinatum"){
    taxon_pft_lookup$pft[i] = "???" # ???
    taxon_pft_lookup$commonName[i] = "vine maple"}

  else if(genusSpecies == "Acer negundo"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "boxelder"}

  else if(genusSpecies == "Acer pensylvanicum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "striped maple"}

  else if(genusSpecies == "Acer rubrum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "red maple"}

  else if(genusSpecies == "Acer saccharinum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "silver maple"}

  else if(genusSpecies == "Acer saccharum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "sugar maple"}

  else if(genusSpecies == "Ailanthus altissima"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "tree of heaven"}

  else if(genusSpecies == "Alnus rubra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"  # deciduous broadleaf tree 
    taxon_pft_lookup$commonName[i] = "red alder"}

  else if(genusSpecies == "Aloysia wrightii"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "Wright's beebrush"}

  else if(genusSpecies == "Amelanchier laevis"){
    taxon_pft_lookup$pft[i] = "???" # ???
    taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}

  else if(genusSpecies == "Amyris elemifera"){
    taxon_pft_lookup$pft[i] = "???"     # ???
    taxon_pft_lookup$commonName[i] = "sea torchwood"}

  else if(genusSpecies == "Artemisia tridentata"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "big sagebrush"}

  else if(genusSpecies == "Atriplex canescens"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "fourwing saltbush"}

  else if(genusSpecies == "Atriplex confertifolia"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$commonName[i] = "shadscale saltbush"}

  else if(genusSpecies == "Betula alleghaniensis"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "yellow birch"}

  else if(genusSpecies == "Betula lenta"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "sweet birch"}

  else if(genusSpecies == "Betula neoalaskana"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "resin birch"}

  else if(genusSpecies == "Betula nigra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "river birch"}

  else if(genusSpecies == "Betula papyrifera"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "paper birch"}

  else if(genusSpecies == "Betula sp."){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$commonName[i] = "birch"}

  else if(genusSpecies == "Bourreria sp."){
    taxon_pft_lookup$pft[i] = "???"   # ???
    taxon_pft_lookup$commonName[i] = "strongbark"}

  else if(genusSpecies == "Bursera simaruba"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"  # "semi-deciduous"
    taxon_pft_lookup$commonName[i] = "gumbo limbo"}

@@

```{r}
kableExtra::kable(taxon_pft_lookup) %>% 
  kableExtra::kable_styling(bootstrap_options = "striped")
```

```{r}
# write the lookup table to a csv 
write_csv(x = taxon_pft_lookup,
          file = here::here("R","output","taxon_pft_lookup.csv"))
```

@@

```{r}
# Set the base url
root <- api("https://plantsdb.xyz")

@@

```{r, eval=FALSE, include=FALSE}
# Specify URL where file is stored 
url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
# Apply download.file function in R
download.file(url, destFile)
```

@@

```{r install_load_packages_bien, message=FALSE, warning=FALSE}
library(BIEN)
library(ape) # Package for working with phylogenies in R
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
```

@@

```{r}
Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")

@@

```{r}
Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
Acer_rubrum_taxonomy[1:3,]
```

@@

>>>>

```{r load_packages, message=FALSE}
library(tidyverse)
library(here)
library(dplyr)
library(kableExtra)
library(neonUtilities)
```

Plant functional types (PFTs) originally proposed: coniferous forest, deciduous forest, woody shrub, grass/herb, and bare ground.

Proposing to update the PFTs based on the MODIS land cover types: Evergreen Needleleaf
Forests, Evergreen Broadleaf Forests, Deciduous Needleleaf Forests, Deciduous Broadleaf
Forests, Shrub.

```{r load_cleaned_spectra, message=FALSE}
# start timer 
start_time <- Sys.time()

# read cleaned spectral library file 
cleaned_spectra <- read_csv(here::here("analysis", "data", "derived_data", "cleaned_spectra.csv"))
```

@@

Run this code chunk to download the NEON woody vegetation structure data: 
```{r download_neon_veg_data, warning=FALSE, message=FALSE, eval=FALSE}
# download woody vegetation structure data for all sites and years 
out_dir <- here::here("analysis","data","raw_data")

@@

```{r read_merge_simplify_growthforms}
# read the vst_apparentindividual table, which has growthForm data
vst_apparentindividual <- read_csv(out_dir <- here::here("analysis","data","raw_data", "filesToStack10098",
                                                         "stackedFiles", "vst_apparentindividual.csv"), 
                                   col_types = cols())

@@

# simplify the growth form into "tree", "shrub", "sapling" etc.  
cleaned_spectra_merged <- cleaned_spectra_merged %>% 
  rowwise() %>%
  # parse out the last word in each entry's growthForm
  # make this code more efficient????? 
  mutate(growthForm_simple = rev(strsplit(growthForm, " ")[[1]])[1])
```

```{r summary_table_taxon_growthforms}
# summarize the merged growthform data
cleaned_spectra_merged %>% 
  distinct(spectraID, .keep_all = TRUE) %>% 
  group_by(genusSpecies, growthForm_simple) %>% 
    tally()

# summarize the merged growthform data with "tree" and "shrub" growth forms
cleaned_spectra_merged %>% 
  distinct(spectraID, .keep_all = TRUE) %>% 
  # keep just the tree and shrub growth forms, remove rows with "sampling" and "liana"
  filter(growthForm_simple %in% c("tree","shrub")) %>% 
  group_by(genusSpecies, growthForm_simple) %>% 
    tally() %>% 
  # display as a nicely formatted table
  kableExtra::kable() %>% 
  kableExtra::kable_styling(bootstrap_options = "striped", 
                            full_width = F, 
                            position = "left")

@@

@@

This code manually adds labels to each taxon (genus and species listed here) for:

- common name
- PFT (based on Macrosystems labels)
- seasonality (evergreen or deciduous)
- leaf type (broadleaf or needleleaf)
- growth form (shrub or forest)

```{r manually_add_pft_labels}
# list unique taxon labels, add an empty columns to fill with plant functional type (PFT) and common name labels.
# PFTs of interest: coniferous_forest, deciduous_forest, woody_shrub, grass_herb, and bare_ground.
taxon_pft_lookup <- data.frame(genusSpecies = unique(cleaned_spectra$genusSpecies),
                               commonName = NA,
                               pft = NA,
                               seasonality = NA,
                               leafType = NA,
                               growthForm = NA) %>% 
  # arrange alphabetically
  dplyr::arrange(genusSpecies)

@@

  if(genusSpecies == "Abies amabilis"){
    taxon_pft_lookup$pft[i] = "coniferous_forest" 
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = "needleleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "Pacific silver fir"}

  else if(genusSpecies == "Abies balsamea"){
    taxon_pft_lookup$pft[i] = "coniferous_forest" 
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = "needleleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "balsam fir"}

  else if(genusSpecies == "Abies lasiocarpa"){
    taxon_pft_lookup$pft[i] = "coniferous_forest"
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = "needleleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "subalpine fir"}

  else if(genusSpecies == "Acacia greggii"){
    taxon_pft_lookup$pft[i] = "woody_shrub" 
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "shrub"
    taxon_pft_lookup$commonName[i] = "catclaw acacia"}

  else if(genusSpecies == "Acer barbatum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "Southern sugar maple"}

  else if(genusSpecies == "Acer circinatum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest" 
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "vine maple"}

  else if(genusSpecies == "Acer negundo"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "boxelder"}

  else if(genusSpecies == "Acer pensylvanicum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "striped maple"}

  else if(genusSpecies == "Acer rubrum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "red maple"}

  else if(genusSpecies == "Acer saccharinum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "silver maple"}

  else if(genusSpecies == "Acer saccharum"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "sugar maple"}

  else if(genusSpecies == "Ailanthus altissima"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "tree of heaven"}

  else if(genusSpecies == "Alnus rubra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "red alder"}

  else if(genusSpecies == "Aloysia wrightii"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "shrub"
    taxon_pft_lookup$commonName[i] = "Wright's beebrush"}

  else if(genusSpecies == "Amelanchier laevis"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "Allegheny serviceberry"}

  else if(genusSpecies == "Amyris elemifera"){
    taxon_pft_lookup$pft[i] = "???" # is not deciduous or coniferous forest
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "sea torchwood"}

  else if(genusSpecies == "Artemisia tridentata"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "shrub"
    taxon_pft_lookup$commonName[i] = "big sagebrush"}

  else if(genusSpecies == "Atriplex canescens"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "shrub"
    taxon_pft_lookup$commonName[i] = "fourwing saltbush"}

  else if(genusSpecies == "Atriplex confertifolia"){
    taxon_pft_lookup$pft[i] = "woody_shrub"
    taxon_pft_lookup$seasonality[i] = "evergreen"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "shrub"
    taxon_pft_lookup$commonName[i] = "shadscale saltbush"}

  else if(genusSpecies == "Betula alleghaniensis"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "yellow birch"}

  else if(genusSpecies == "Betula lenta"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "sweet birch"}

  else if(genusSpecies == "Betula neoalaskana"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "resin birch"}

  else if(genusSpecies == "Betula nigra"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "river birch"}

  else if(genusSpecies == "Betula papyrifera"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "paper birch"}

  else if(genusSpecies == "Betula sp."){
    taxon_pft_lookup$pft[i] = "deciduous_forest"
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = "broadleaf"
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "birch"}

  else if(genusSpecies == "Bourreria sp."){
    taxon_pft_lookup$pft[i] = "???"   
    taxon_pft_lookup$seasonality[i] = ""
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "strongbark"}

  else if(genusSpecies == "Bursera simaruba"){
    taxon_pft_lookup$pft[i] = "deciduous_forest"  
    taxon_pft_lookup$seasonality[i] = "deciduous"
    taxon_pft_lookup$leafType[i] = ""
    taxon_pft_lookup$growthForm[i] = "forest"
    taxon_pft_lookup$commonName[i] = "gumbo limbo"}

@@

```{r taxon_pft_table}
kableExtra::kable(taxon_pft_lookup) %>% 
  kableExtra::kable_styling(bootstrap_options = "striped")
```

```{r write_csv}
# write the lookup table to a csv 
write_csv(x = taxon_pft_lookup,
          file = here::here("R","output","taxon_pft_lookup.csv"))
```

@@

```{r usda_plants}
# Set the base url
root <- api("https://plantsdb.xyz")

@@

```{r try_download, eval=FALSE, include=FALSE}
# Specify URL where file is stored 
url <- "https://uofi.app.box.com/v/usdaplants/folder/6688524249/usda_plants_20160713.csv"
destFile <- file.path(here::here(), "analysis","data", "usda_plants_20160713.csv")
# Apply download.file function in R
download.file(url, destFile)
```

@@

```{r load_packages_bien, message=FALSE, warning=FALSE}
library(BIEN)
library(ape) # Package for working with phylogenies in R
library(maps) # Useful for making quick maps of occurrences
library(sp) # A package for spatial data
```

@@

```{r bien_taxonomy_1}
Abies_taxonomy<-BIEN_taxonomy_genus(genus = "Abies")

@@

```{r bien_taxonomy_2}
Acer_rubrum_taxonomy <- BIEN_taxonomy_species(species = "Acer rubrum")
Acer_rubrum_taxonomy[1:3,]
```

@@

Time the code. How long did this take? 
```{r how_long_did_this_take}
# stop timer
end_time <- Sys.time()

elapsed_time <- end_time - start_time

print(difftime(end_time, start_time, 
         units = c("auto", "secs", "mins", "hours",
                   "days", "weeks")))
```
